##Prediction Assignment Writeup

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har

For the prediction of how welll individuals performed the assigned exercise six young health participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions: exactly according to the specification (Class A), throwing the elbows to the front (Class B), lifting the dumbbell only halfway (Class C), lowering the dumbbell only halfway (Class D) and throwing the hips to the front (Class E).

This report aims to use machine learning algoritmhs to predict the class of exercise the individuals was performing by using meaurements available from devices such as Jawbone Up, Nike FuelBand, and Fitbit.

##Data Cleaning
The data for this project come was obteined from http://groupware.les.inf.puc-rio.br/har. Two data set were available a training set and a test set for which 20 individuals without any classification for the class of exercise was available.


```{r}
#Data loading
pmlTrain<-read.csv("pml-training.csv", header=T, na.strings=c("NA", "#DIV/0!"))
pmlTest<-read.csv("pml-testing.csv", header=T, na.string=c("NA", "#DIV/0!"))
```


Training data was partitioned and preprocessed using the code described below. In brief, all variables with at least one "NA" were excluded from the analysis. Variables related to time and user information were excluded for a total of 51 variables and 19622 class measurements. Same variables were mainteined in the test data set (Validation dataset) to be used for predicting the 20 test cases provided.


```{r}
## NA exclusion for all available variables
noNApmlTrain<-pmlTrain[, apply(pmlTrain, 2, function(x) !any(is.na(x)))] 
dim(noNApmlTrain)
```


```{r}
# variables with user information, time and undefined
cleanpmlTrain<-noNApmlTrain[,-c(1:8)]
dim(cleanpmlTrain)
```

```{r}
# 20 test cases provided clean info - Validation data set
cleanpmltest<-pmlTest[,names(cleanpmlTrain[,-52])]
dim(cleanpmltest)
```

##Data Partitioning and Prediction Process
The cleaned downloaded data set was subset in order to generate a test set independent from the 20 cases provided set. Partitioning was performed to obtain a 75% training set and a 25% test set.


```{r}
#data cleaning
library(caret)
inTrain<-createDataPartition(y=cleanpmlTrain$classe, p=0.75,list=F)
training<-cleanpmlTrain[inTrain,] 
test<-cleanpmlTrain[-inTrain,] 
```



```{r}
#Training and test set dimensions
dim(training)
```

```{r}
dim(test)
```


##Results and Conclusions
Random forest trees were generated for the training dataset using cross-validation. Then the generated algorithm was examnined under the partitioned training set to examine the accuracy and estimated error of prediction. By using 51 predictors for five classes using cross-validation at a 5-fold an accuracy of 99.2% with a 95% CI [0.989-0.994] was achieved accompanied by a Kappa value of 0.99.
```{r}
library(caret)
library(randomForest)
library(e1071)
set.seed(13333)
fitControl2<-trainControl(method="cv", number=5, allowParallel=T, verbose=T)
rffit<-train(classe~.,data=training, method="rf", trControl=fitControl2, verbose=F)
```


```{r}
predrf<-predict(rffit, newdata=test)
confusionMatrix(predrf, test$classe)
pred20<-predict(rffit, newdata=cleanpmltest)
```



```{r}
# Output for the prediction of the 20 cases provided
pred20
```


A boosting algorithm was also run to confirm and be able to compare predictions. Data is not shown but the boosting approach presented less accuracy (96%) (Data not shown). However, when the predictions for the 20 test cases were compared match was same for both ran algorimths.
```{r}
library(gbm)
fitControl2<-trainControl(method="cv", number=5, allowParallel=T, verbose=T)
gmbfit<-train(classe~.,data=training, method="gbm", trControl=fitControl2, verbose=F)
gmbfit$finalModel
class(gmbfit)
predgmb<-predict(gmbfit, newdata=test)
confusionMatrix(predgmb, test$classe)
predtrain<-predict(gmbfit, newdata=training)
confusionMatrix(predtrain, training$classe)
predtrain<-predict(gmbfit, newdata=training)
confusionMatrix(predtrain, training$classe)
```


Once, the predictions were obtained for the 20 test cases provided, the below shown script was used to obtain single text files to be uploaded to the courses web site to comply with the submission assigment. 20 out of 20 hits also confirmed the accuracy of the obtained models.
```{r}

pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
pml_write_files(pred20)
```

